---
title: "LSAS-SR Confirmatory and Exploratory Factor Analyses"
author: "NÖ"
date: "2026-07-23"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
```

## Packages and data

```{r packages-data}
library(dplyr)
library(here)
library(knitr)
library(lavaan)
library(psych)
library(readr)
library(semTools)
library(tidyr)

smarvus_dat <- here("data", "smarvus_complete_290723.csv") |>
  read_csv(
    show_col_types = FALSE,
    guess_max = Inf
  )
```

## Exclusions and preparation

```{r exclusions}
lsas_items <- paste0("Q12.1_", 1:24)

smarvus_dat_a <- smarvus_dat |>
  filter(attention_amnesty == "Yes")

smarvus_dat_b <- smarvus_dat_a |>
  filter(
    Q7.1_24 == "1",
    Q8.1_21 == "5",
    Q11.1_9 == "3",
    Q9.1_22 == "1",
    Q13.1_17 == "2",
    Q15.1_9 == "4"
  )

smarvus_dat_all <- smarvus_dat_b |>
  drop_na(all_of(lsas_items)) |>
  mutate(across(all_of(lsas_items), as.numeric))

exclusion_summary <- tibble(
  stage = c(
    "Initial sample",
    "Attention amnesty",
    "Attention checks",
    "Complete LSAS-SR data"
  ),
  n = c(
    nrow(smarvus_dat),
    nrow(smarvus_dat_a),
    nrow(smarvus_dat_b),
    nrow(smarvus_dat_all)
  ),
  excluded_at_stage = c(
    NA_integer_,
    nrow(smarvus_dat) - nrow(smarvus_dat_a),
    nrow(smarvus_dat_a) - nrow(smarvus_dat_b),
    nrow(smarvus_dat_b) - nrow(smarvus_dat_all)
  )
)

kable(exclusion_summary)

# Confirm that parsing issues do not affect variables used in the CFA.
parsing_issues <- problems(smarvus_dat)
analysis_variables <- c(
  lsas_items,
  "attention_amnesty",
  "Q7.1_24", "Q8.1_21", "Q11.1_9",
  "Q9.1_22", "Q13.1_17", "Q15.1_9"
)

parsing_issues_analysis_variables <- parsing_issues |>
  mutate(variable = names(smarvus_dat)[col]) |>
  filter(variable %in% analysis_variables)

stopifnot(nrow(parsing_issues_analysis_variables) == 0L)

# Check that every item has four ordered response categories.
lapply(smarvus_dat_all[lsas_items], table, useNA = "ifany")
```

## Demographic summaries

```{r demographics}
demographic_overview <- tibble(
  characteristic = c(
    "Final sample",
    "Languages",
    "Countries",
    "Universities"
  ),
  value = c(
    nrow(smarvus_dat_all),
    n_distinct(smarvus_dat_all$language, na.rm = TRUE),
    n_distinct(smarvus_dat_all$country, na.rm = TRUE),
    n_distinct(smarvus_dat_all$university, na.rm = TRUE)
  )
)

summarise_demographic <- function(data, variable) {
  data |>
    count({{ variable }}, name = "n", .drop = FALSE) |>
    mutate(percentage = round(100 * n / sum(n), 2)) |>
    arrange(desc(n))
}

language_demographics <- summarise_demographic(smarvus_dat_all, language)
country_demographics <- summarise_demographic(smarvus_dat_all, country)
university_demographics <- summarise_demographic(smarvus_dat_all, university)
gender_demographics <- summarise_demographic(smarvus_dat_all, gender)
age_demographics <- summarise_demographic(smarvus_dat_all, age)

kable(demographic_overview)
kable(language_demographics)
kable(country_demographics)
kable(university_demographics)
kable(gender_demographics)
kable(age_demographics)
```

## Descriptive statistics using the original 0--3 scoring

The Qualtrics responses are coded 1--4. Subtracting one is needed for score
descriptives.

```{r descriptives}
lsas_scored <- smarvus_dat_all |>
  transmute(across(all_of(lsas_items), ~ .x - 1))

item_descriptives <- tibble(
  item = lsas_items,
  mean = vapply(lsas_scored, mean, numeric(1), na.rm = TRUE),
  sd = vapply(lsas_scored, sd, numeric(1), na.rm = TRUE),
  minimum = vapply(lsas_scored, min, numeric(1), na.rm = TRUE),
  maximum = vapply(lsas_scored, max, numeric(1), na.rm = TRUE)
) |>
  mutate(across(where(is.numeric), ~ round(.x, 2)))

kable(item_descriptives)
```

## Model specifications

```{r models}
liebowitz_model <- '
Performance =~ Q12.1_1 + Q12.1_2 + Q12.1_3 + Q12.1_4 +
               Q12.1_6 + Q12.1_8 + Q12.1_9 + Q12.1_13 +
               Q12.1_14 + Q12.1_16 + Q12.1_17 + Q12.1_20 +
               Q12.1_21

SocialInteraction =~ Q12.1_5 + Q12.1_7 + Q12.1_10 + Q12.1_11 +
                     Q12.1_12 + Q12.1_15 + Q12.1_18 + Q12.1_19 +
                     Q12.1_22 + Q12.1_23 + Q12.1_24
'
safren_model <- '
SocialInteraction =~ Q12.1_5 + Q12.1_7 + Q12.1_10 + Q12.1_11 +
                     Q12.1_12 + Q12.1_18 + Q12.1_19 + Q12.1_21
PublicSpeaking =~ Q12.1_2 + Q12.1_6 + Q12.1_15 + Q12.1_16 + Q12.1_20
Observation =~ Q12.1_1 + Q12.1_9 + Q12.1_13 + Q12.1_17
EatNDrink =~ Q12.1_3 + Q12.1_4
'

oakman_model <- '
SocialInteraction =~ Q12.1_5 + Q12.1_7 + Q12.1_10 + Q12.1_11 +
                     Q12.1_12 + Q12.1_14 + Q12.1_18 + Q12.1_19 +
                     Q12.1_21 + Q12.1_22 + Q12.1_23 + Q12.1_24
PublicSpeaking =~ Q12.1_2 + Q12.1_6 + Q12.1_15 + Q12.1_16 + Q12.1_20
Observation =~ Q12.1_1 + Q12.1_8 + Q12.1_9 + Q12.1_13 + Q12.1_17
EatNDrink =~ Q12.1_3 + Q12.1_4
'

baker_model <- '
SocialInteraction =~ Q12.1_12 + Q12.1_11 + Q12.1_10 + Q12.1_7 +
                     Q12.1_5 + Q12.1_19 + Q12.1_18 + Q12.1_21 + Q12.1_14
NonVerbPerformance =~ Q12.1_17 + Q12.1_9 + Q12.1_8
Ingestion =~ Q12.1_3 + Q12.1_4
PublicPerformance =~ Q12.1_20 + Q12.1_6 + Q12.1_16 + Q12.1_15
Assertiveness =~ Q12.1_13 + Q12.1_22 + Q12.1_24 + Q12.1_1
'
```

## Fit all models using WLSMV

The explicit `delta` parameterization is lavaan's default for ordered outcomes.
It is stated here so that subsequent measurement-invariance analyses can use
exactly the same settings.

```{r fit-models}
get_model_items <- function(model_syntax) {
  lavNames(lavaanify(model_syntax), type = "ov")
}

fit_wlsmv <- function(model_syntax, data) {
  cfa(
    model = model_syntax,
    data = data,
    ordered = get_model_items(model_syntax),
    estimator = "WLSMV",
    parameterization = "delta",
    std.lv = TRUE
  )
}

liebowitz_fit <- fit_wlsmv(liebowitz_model, smarvus_dat_all)
safren_fit <- fit_wlsmv(safren_model, smarvus_dat_all)
oakman_fit <- fit_wlsmv(oakman_model, smarvus_dat_all)
baker_fit <- fit_wlsmv(baker_model, smarvus_dat_all)

fits <- list(
  "Liebowitz" = liebowitz_fit,
  "Safren" = safren_fit,
  "Oakman" = oakman_fit,
  "Baker" = baker_fit
)

convergence_table <- tibble(
  model = names(fits),
  converged = vapply(fits, lavInspect, logical(1), what = "converged"),
  observed_items = vapply(fits, function(x) length(lavNames(x, type = "ov")), integer(1))
)

kable(convergence_table)
```

## Model-fit results

WLSMV does not provide likelihood-based AIC or BIC. Models are therefore
evaluated using robust CFI, robust TLI, robust RMSEA and SRMR. Because Safren
and Baker omit some items, their fit indices are not strictly comparable with
those of the 24-item models.

```{r fit-table}
get_fit_value <- function(fit_values, preferred, fallback = NULL) {
  if (preferred %in% names(fit_values) && !is.na(fit_values[[preferred]])) {
    return(unname(fit_values[[preferred]]))
  }
  if (!is.null(fallback) && fallback %in% names(fit_values)) {
    return(unname(fit_values[[fallback]]))
  }
  NA_real_
}

extract_fit <- function(fit, model_name) {
  fm <- fitMeasures(fit)
  tibble(
    model = model_name,
    n_items = length(lavNames(fit, type = "ov")),
    chisq = get_fit_value(fm, "chisq.scaled", "chisq"),
    df = get_fit_value(fm, "df.scaled", "df"),
    p = get_fit_value(fm, "pvalue.scaled", "pvalue"),
    cfi = get_fit_value(fm, "cfi.robust", "cfi.scaled"),
    tli = get_fit_value(fm, "tli.robust", "tli.scaled"),
    rmsea = get_fit_value(fm, "rmsea.robust", "rmsea.scaled"),
    rmsea_lower = get_fit_value(fm, "rmsea.ci.lower.robust", "rmsea.ci.lower.scaled"),
    rmsea_upper = get_fit_value(fm, "rmsea.ci.upper.robust", "rmsea.ci.upper.scaled"),
    srmr = get_fit_value(fm, "srmr")
  )
}

cfa_fit_table <- bind_rows(Map(extract_fit, fits, names(fits)))

cfa_fit_table_report <- cfa_fit_table |>
  mutate(
    chisq = round(chisq, 2),
    df = round(df),
    p = ifelse(p < .001, "< .001", sprintf("%.3f", p)),
    across(c(cfi, tli, rmsea, rmsea_lower, rmsea_upper, srmr), ~ round(.x, 3))
  )

kable(cfa_fit_table_report)
```

## Standardized loadings

```{r loadings}
extract_loadings <- function(fit, model_name) {
  standardizedSolution(fit) |>
    filter(op == "=~") |>
    transmute(
      model = model_name,
      factor = lhs,
      item = rhs,
      loading = est.std,
      se = se,
      z = z,
      p = pvalue
    )
}

cfa_loadings <- bind_rows(Map(extract_loadings, fits, names(fits)))

kable(
  cfa_loadings |>
    mutate(across(c(loading, se, z), ~ round(.x, 3)))
)
```

## Factor correlations

```{r factor-correlations}
extract_factor_correlations <- function(fit, model_name) {
  latent_names <- lavNames(fit, type = "lv")
  standardizedSolution(fit) |>
    filter(
      op == "~~",
      lhs != rhs,
      lhs %in% latent_names,
      rhs %in% latent_names
    ) |>
    transmute(
      model = model_name,
      factor_1 = lhs,
      factor_2 = rhs,
      correlation = est.std
    )
}

factor_correlations <- bind_rows(
  Map(extract_factor_correlations, fits, names(fits))
)

kable(
  factor_correlations |>
    mutate(correlation = round(correlation, 3))
)
```

## Model-based reliability

```{r reliability}
## Model-based reliability and Cronbach's alpha

stopifnot(packageVersion("semTools") >= "0.5.9")

# Model-based omega on the observed ordinal response scale
extract_omega <- function(fit, model_name) {

  result <- semTools::compRelSEM(
    fit,
    tau.eq = FALSE,
    ord.scale = TRUE,
    obs.var = TRUE
  )

  tibble(
    model = model_name,
    factor = names(result),
    omega = vapply(
      result,
      function(x) as.numeric(x)[1],
      numeric(1)
    )
  )
}

omega_results <- bind_rows(
  Map(
    extract_omega,
    fits,
    names(fits)
  )
)

# Classical Cronbach's alpha calculated from observed item scores
cronbach_alpha <- function(data) {
  item_covariance <- cov(data, use = "complete.obs")
  number_of_items <- ncol(data)

  number_of_items / (number_of_items - 1) *
    (1 - sum(diag(item_covariance)) / sum(item_covariance))
}

extract_alpha <- function(fit, model_name, data) {

  loading_structure <- parameterEstimates(fit) |>
    filter(op == "=~") |>
    select(factor = lhs, item = rhs)

  bind_rows(
    lapply(unique(loading_structure$factor), function(factor_name) {

      factor_items <- loading_structure |>
        filter(factor == factor_name) |>
        pull(item)

      tibble(
        model = model_name,
        factor = factor_name,
        alpha = cronbach_alpha(data[factor_items])
      )
    })
  )
}

alpha_results <- bind_rows(
  Map(
    function(fit, model_name) {
      extract_alpha(
        fit,
        model_name,
        smarvus_dat_all
      )
    },
    fits,
    names(fits)
  )
)

reliability_table <- full_join(
  omega_results,
  alpha_results,
  by = c("model", "factor")
)

reliability_table_report <- reliability_table |>
  mutate(
    omega = round(omega, 3),
    alpha = round(alpha, 3)
  )

kable(reliability_table_report)
```

## Full summaries and diagnostics

```{r summaries}
lapply(
  fits,
  summary,
  fit.measures = TRUE,
  standardized = TRUE
)

# Check residual variances and standardized solutions for inadmissible values.
lapply(fits, lavInspect, what = "post.check")
```

## Save reproducible results

```{r save-results}
results_directory <- here("results")
dir.create(results_directory, showWarnings = FALSE)

write_csv(exclusion_summary, file.path(results_directory, "exclusion_summary.csv"))
write_csv(demographic_overview, file.path(results_directory, "demographic_overview.csv"))
write_csv(language_demographics, file.path(results_directory, "demographics_language.csv"))
write_csv(country_demographics, file.path(results_directory, "demographics_country.csv"))
write_csv(university_demographics, file.path(results_directory, "demographics_university.csv"))
write_csv(gender_demographics, file.path(results_directory, "demographics_gender.csv"))
write_csv(age_demographics, file.path(results_directory, "demographics_age.csv"))
write_csv(cfa_fit_table, file.path(results_directory, "WLSMV_CFA_fit_indices.csv"))
write_csv(cfa_loadings, file.path(results_directory, "WLSMV_CFA_factor_loadings.csv"))
write_csv(factor_correlations, file.path(results_directory, "WLSMV_factor_correlations.csv"))
write_csv(item_descriptives, file.path(results_directory, "LSAS_item_descriptives.csv"))
write_csv(
  reliability_table,
  file.path(results_directory, "WLSMV_model_based_reliability.csv")
)

saveRDS(fits, file.path(results_directory, "WLSMV_CFA_models.rds"))

capture.output(
  sessionInfo(),
  file = file.path(results_directory, "sessionInfo.txt")
)
```

## Model-comparison note

A formal chi-square difference test was not conducted because the four CFA models were non-nested and did not all include the same observed indicators. The Liebowitz and Oakman models included all 24 LSAS-SR items, whereas the Safren and Baker models included 19 and 22 items, respectively. Each model was therefore evaluated primarily in terms of its absolute fit, and comparisons of fit indices across models were interpreted descriptively. WLSMV chi-square difference testing is appropriate only for nested models fitted to the same observed indicators using the same estimator and ordered-categorical specification.

## Exploratory factor analysis

The EFA was based on polychoric correlations to account for the ordinal
response format of the LSAS-SR items. The analyses include the preliminary three-factor analysis, the two
VSS calls, the factor and component parallel analyses based on the precomputed
polychoric correlation matrix, the later ML parallel analysis, and the final
eight-factor EFA. The final EFA used the `psych::fa()` defaults of
minres extraction and oblimin rotation, with ten Berge factor scores.

```{r efa-data-suitability}
efa_data <- smarvus_dat_all |>
  dplyr::select(dplyr::all_of(lsas_items))

n_efa <- nrow(efa_data)

# Calculate the polychoric correlation matrix once.
efa_poly_result <- psych::polychoric(efa_data)
efa_poly <- efa_poly_result$rho

# Assess whether the correlation matrix is suitable for factor analysis.
efa_kmo <- psych::KMO(efa_poly)

efa_bartlett <- psych::cortest.bartlett(
  efa_poly,
  n = n_efa
)

efa_suitability <- tibble(
  overall_kmo = unname(efa_kmo$MSA),
  minimum_item_kmo = unname(min(efa_kmo$MSAi)),
  maximum_item_kmo = unname(max(efa_kmo$MSAi)),
  bartlett_chisq = unname(efa_bartlett$chisq),
  bartlett_df = unname(efa_bartlett$df),
  bartlett_p = unname(efa_bartlett$p.value)
)

kable(
  efa_suitability |>
    mutate(across(where(is.numeric), ~ round(.x, 3)))
)
```

```{r efa-eigenvalues-scree}
# Correlation-matrix eigenvalues are used for the scree plot.
efa_eigenvalues <- eigen(efa_poly)$values

efa_eigenvalue_table <- tibble(
  factor = seq_along(efa_eigenvalues),
  eigenvalue = efa_eigenvalues
)

kable(
  efa_eigenvalue_table |>
    mutate(eigenvalue = round(eigenvalue, 3))
)

plot(
  seq_along(efa_eigenvalues),
  efa_eigenvalues,
  type = "b",
  pch = 19,
  xlab = "Factor",
  ylab = "Eigenvalue",
  main = "Scree Plot of LSAS-SR Polychoric Correlations"
)

```

```{r efa-ad-original-approach}
# Recreate the object names used in the original code.
cor_dat <- efa_data
raq_poly <- psych::polychoric(cor_dat)
raq_cor <- raq_poly$rho
just_lsas <- smarvus_dat_all |>
  dplyr::select(dplyr::all_of(lsas_items))

# Preliminary three-factor EFA
raq_fa <- psych::fa(
  cor_dat,
  nfactors = 3,
  scores = "tenBerge",
  cor = "poly"
)

raq_fa

# First ML-based VSS analysis
efa_output <- psych::vss(
  just_lsas,
  fa = "fa",
  fm = "ml",
  cor = "poly",
  SMC = TRUE
)

efa_output

# Factor and component parallel analyses based on the precomputed
# polychoric correlation matrix.
set.seed(20260724)

efa_parallel_factor <- psych::fa.parallel(
  raq_cor,
  n.obs = 8338,
  fa = "fa",
  cor = "poly"
)

set.seed(20260724)

efa_parallel_component <- psych::fa.parallel(
  raq_cor,
  n.obs = 8338,
  fa = "pc",
  cor = "poly"
)

# Later ML parallel analysis and VSS analysis
efa_dat_all <- dplyr::select(
  smarvus_dat_all,
  dplyr::all_of(lsas_items)
)

set.seed(20260724)

efa_out_all_1 <- psych::fa.parallel(
  efa_dat_all,
  fa = "fa",
  fm = "ml",
  cor = "poly",
  SMC = TRUE,
  n.iter = 100
)

efa_out_all_2 <- psych::vss(
  efa_dat_all,
  fm = "ml",
  cor = "poly"
)

efa_out_all_1
efa_out_all_2

efa_vss_original_table <- efa_out_all_2$vss.stats |>
  as.data.frame() |>
  tibble::rownames_to_column("n_factors")

kable(
  efa_vss_original_table |>
    dplyr::select(
      n_factors,
      dof,
      chisq,
      RMSEA,
      BIC,
      SABIC,
      SRMR
    ) |>
    dplyr::mutate(
      n_factors = as.integer(n_factors),
      dplyr::across(where(is.numeric), ~ round(.x, 3))
    )
)

efa_original_retention_summary <- tibble(
  analysis = c(
    "Factor parallel analysis on raq_cor",
    "Component parallel analysis on raq_cor",
    "ML factor parallel analysis (SMC = TRUE)",
    "ML VSS: lowest BIC among 1-8 factors",
    "ML VSS: lowest SABIC among 1-8 factors"
  ),
  suggested_dimensions = c(
    efa_parallel_factor$nfact,
    efa_parallel_component$ncomp,
    efa_out_all_1$nfact,
    which.min(efa_vss_original_table$BIC),
    which.min(efa_vss_original_table$SABIC)
  )
)

kable(efa_original_retention_summary)
```

```{r efa-final-model}
# Final EFA call. Because fm and rotate were not supplied,
# psych::fa() used minres extraction and oblimin rotation.
efa_fit <- psych::fa(
  cor_dat,
  nfactors = 8,
  scores = "tenBerge",
  cor = "poly"
)

efa_fit

print(
  efa_fit$loadings,
  cutoff = 0,
  digits = 3
)

efa_variance_explained <- as.data.frame(efa_fit$Vaccounted) |>
  tibble::rownames_to_column("statistic")

kable(
  efa_variance_explained |>
    mutate(across(where(is.numeric), ~ round(.x, 3)))
)
```

## EFA loading summary

Salient loadings are defined using the unrounded loading estimates and a
threshold of \(\lvert\lambda\rvert \ge .32\). Applying the threshold before
rounding avoids classifying estimates slightly below .32 as salient merely
because they round to .32.

```{r efa-loading-summary}
efa_loading_matrix <- unclass(efa_fit$loadings)

efa_loadings <- as.data.frame(efa_loading_matrix) |>
  tibble::rownames_to_column("item")

efa_primary_loadings <- tibble(
  item = rownames(efa_loading_matrix),
  primary_factor = colnames(efa_loading_matrix)[
    max.col(abs(efa_loading_matrix), ties.method = "first")
  ],
  primary_loading = efa_loading_matrix[
    cbind(
      seq_len(nrow(efa_loading_matrix)),
      max.col(abs(efa_loading_matrix), ties.method = "first")
    )
  ]
) |>
  mutate(
    absolute_loading = abs(primary_loading),
    salient = absolute_loading >= .32
  )

kable(
  efa_primary_loadings |>
    mutate(
      primary_loading = round(primary_loading, 3),
      absolute_loading = round(absolute_loading, 3)
    )
)

efa_cross_loadings <- tibble(
  item = rownames(efa_loading_matrix),
  number_salient_loadings = rowSums(abs(efa_loading_matrix) >= .32),
  cross_loading = number_salient_loadings > 1
)

kable(efa_cross_loadings)

# Items with no salient loading.
efa_primary_loadings |>
  filter(!salient)

# Items assigned to factors using their largest absolute salient loading.
efa_primary_loadings |>
  filter(salient) |>
  arrange(primary_factor, desc(absolute_loading))

# Items with salient loadings on more than one factor.
efa_cross_loadings |>
  filter(cross_loading)

efa_factor_correlations <- as.data.frame(efa_fit$Phi) |>
  tibble::rownames_to_column("factor")

kable(
  efa_factor_correlations |>
    mutate(across(where(is.numeric), ~ round(.x, 3)))
)
```

## Save EFA results

```{r save-efa-results}
write_csv(
  efa_suitability,
  file.path(results_directory, "EFA_data_suitability.csv")
)

write_csv(
  efa_eigenvalue_table,
  file.path(results_directory, "EFA_correlation_eigenvalues.csv")
)

write_csv(
  efa_original_retention_summary,
  file.path(results_directory, "EFA_AD_retention_summary.csv")
)

write_csv(
  efa_variance_explained,
  file.path(results_directory, "EFA_variance_explained.csv")
)

write_csv(
  efa_loadings,
  file.path(results_directory, "EFA_factor_loadings.csv")
)

write_csv(
  efa_primary_loadings,
  file.path(results_directory, "EFA_primary_loadings.csv")
)

write_csv(
  efa_cross_loadings,
  file.path(results_directory, "EFA_cross_loadings.csv")
)

write_csv(
  efa_factor_correlations,
  file.path(results_directory, "EFA_factor_correlations.csv")
)

saveRDS(
  list(
    preliminary_three_factor_efa = raq_fa,
    first_vss_ml = efa_output,
    parallel_analysis_factor = efa_parallel_factor,
    parallel_analysis_component = efa_parallel_component,
    parallel_analysis_ml = efa_out_all_1,
    second_vss_ml = efa_out_all_2,
    final_eight_factor_efa = efa_fit
  ),
  file.path(results_directory, "EFA_models.rds")
)

# Refresh the reproducibility record after all CFA and EFA analyses.
capture.output(
  sessionInfo(),
  file = file.path(results_directory, "sessionInfo.txt")
)
```
